Peptide retention time prediction

### 2015-11-11 retention.tex
Antecedents to this etc.

Today a majority of the mass spectrometry (MS) protocols in use have a
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms for data interpretation a major
area within proteomics research. A particularly difficult challenge in
this context is posed by the tremendous complexity of the data.
Proteins display a wide range of physico-chemical properties and are
often found at very different amounts in complex samples, spanning as
much as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications~\cite{Lemeer2009}. To manage this
diversity, increasing efforts are made to extract additional
information besides the fragmentation spectra from each MS experiment.

An example of such additional information is represented by retention
times, {\em i.e.} the time points when peptides elute from the liquid
chromatography column.  The main purpose of liquid chromatography is
to reduce the number of peptides simultaneously injected into the mass
spectrometer. This increases the probability of each individual
peptide to be selected for fragmentation, which in its turn, increases
the proteome coverage achieved in an experiment.  However, besides
this, the separation process in itself provides useful information
about the analytes. On the one hand, the instrument always records the
time when a peptide elutes from the chromatography column, {\em i.e.}
the retention time. On the other hand, the retention time can also be
estimated by examining the sequence of the peptide. The availability
of both predicted and observed retention times has proven to be a
powerful feature, with several applications in MS-based proteomics
\cite{Palmblad2013}.

@@

At the moment, Reversed-Phase Liquid Chromatography (RPLC) is by far
the preferred method for separating peptides prior to mass
spectrometry analysis \cite{Xie201278}. This is mainly because the
method provides good resolution, gives reproducible results, and most
of all, is compatible with electrospray ionization. In virtually all
modern mass spectrometers the RPLC system can be easily connected to
the ionization source, such that the peptides can be automatically
transferred to mass spectrometry analysis as they elute from the
chromatography column.

@@

% say something about gradient and washing 
Since the proportion of solvent B determines the elution of peptides,
the scheme used to add solvent B is an essential aspect of any RPLC
experiment. Currently, the most commonly used approach is based on
{\em linear gradients}, where the increase in solvent B follows a
linear function. Most of the peptides elute before the percentage of
solvent B reaches 40\%, and thus gradients often go from 2 to 40\%
B. A high increase in percentage B at the end of the run ensures that
all the remaining compounds are washed off from the column.

@@

Liquid chromatography gives reproducible results. Except for small
shifts from run to run, the same peptide elutes at about the same time
when a sample is reanalyzed under identical experimental conditions
(Figure~\ref{fig:repr}). This suggests that peptides engage in
well-defined interactions during separation, which depend only on the
experimental setup employed and the properties of the peptides
themselves. However, the properties of the peptides are determined to
a large extent by their amino-acid sequence. As a consequence, given a
chromatography setup, the sequence of a peptide should provide
virtually all the information needed to predict its retention time.

@@

% The first method 
The first method to implement this idea used a set of 25 short
peptides together with their observed retention times to derive
retention coefficients for each of the amino acids and end groups
present in the sequences \cite{meek1980}. The coefficients were
calculated using a form of iterative regression and were estimated
using only the amino-acid composition of the peptides, without any
information about the position of each amino acid in the
sequence. However, since shuffled versions of the same peptide
sequence had previously been separated on a RPLC column, the author
pointed out that additional parameters would be required for
predicting the retention time of longer peptides.

@@

% It became available that additional information is needed; anc oclumn is important  examples 
Indeed, a subsequent study showed that even for short peptides,
distinct retention coefficients are obtained when an amino acid is
located at different positions in a sequence \cite{Houghten1987}. In
this context, the authors speculated that different retention
coefficients would be needed for each position in a peptide
sequence. Although the exact chemical mechanisms underlying this
observation are still not fully understood, a number of studies have
pinpointed individual factors that alter the retention coefficients of
amino acids. As an example, it has been shown that the retention
contribution of an amino acid is affected by the overall
hydrophobicity of the peptide \cite{Mant2006}, the number of positive
charges \cite{Mant2006} or presence of elements of secondary
structure \cite{Zhou1990}. In addition, modifications in
chromatography setup such as changing the composition of the mobile or
stationary phases also modify the elution pattern of the
peptides \cite{Browne1982, Guo1987, Gilar2010}.

@@

A limitation of index-based methods is that they are optimized for
predicting retention times for certain chromatography setups. Thus,
their prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}. Consequently, more advanced methods such as {\sc
  Elude} \cite{elude1,elude2} estimate retention time indices for each
dataset as hand. In these cases the indices are used to generate
features that are used as an input for final retention time
estimates. We will come back to this in section \ref{sec:ML}.

% - sum of retention coefficients (Meek reference 22) - different RT
% coefficients and the observation that this coefficients depend on
% the chromatography system (Browne ref 24) - investogate the
% contribution of individual amino acids (Guo 26,27) - all the work
% above was done on synthetic peptides os simple samples. Then RT
% prediction for complex sa% mples was proposed by (Palmblad 28-30) -
% important the exact position of each aa (pref 27) - additional rules
% included peptide length (ref 32), position of K and R residues [ref
% 33], and in the e% nd that the position of each amino acid at each
% position is important [ref 34] - amphipathicity [ref 35] - SSRCalc
% [ref 36-39]

@@

The method was extended by representing each peptide by 1052 instead
of 20 features in a follow-up study~\cite{petritis2006improved}. The
new descriptors encoded the identity of each amino acid at each
position in a sequence, as well as the length and the hydrophobic
moment of the peptide. The new neural network was trained on more than
300,000 peptide sequences. To obtain such a large number of peptides
with known retention times, a strategy to normalize elution times was
developed. This allowed pooling of peptides identified in more than
12,000 shotgun experiments. The trained algorithm reported better
performance for the 1303 peptides on which it was evaluated, than its
predecessor.

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer {\em et al.} proposed a predictor based on support vector
machines (SVMs) \cite{klammer2007improving}. Their method required
fewer training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction accuracy.
Subsequently, a more elaborate SVM-based method, {\sc RTPredict}, that
uses a customized kernel function was proposed \cite{rtpredict,
  rtpredictImproved}. This approach provided an algorithm that
required as little as 50 peptides for training.

@@

Moruz {\em et al.}~\cite{elude1} developed a software package for
retention time prediction named {\sc Elude}. One of the main novelty
points of {\sc Elude} was the calculation of peptide features based on
both a general hydrophobicity index, as well as a retention index
trained for each dataset at hand. The features included properties
such as average hydrophobicity, the N- and C-termini's hydrophobicity,
the stretch of the five most and least hydrophobic amino acids,
helicity, and the number of occurrences of the 25\% most or least
hydrophobic amino acids. This ensured that the algorithm could be
applied to different chromatography setups with minimal loss in
performance. For cases when little training data was available, {\sc
Elude} implemented an additional workflow based on a library of
pre-trained retention models. This workflow required only a few tens
of training peptides to select and calibrate the most suitable model
from the library. Both the selection and calibration were performed
using statistical methods robust to outliers.

@@

To address this, in Moruz {\em et al.} \cite{elude2} extended the
first version of their algorithm {\sc ELUDE} to encompass modified
peptides as well. This was straightforward to achieve once the authors
observed that many of the predictive features of their method were all
functions of the peptide's amino acid sequence and a retention
index. As such retention indices are estimated for each dataset at
hand, at the beginning of the training procedure, they can trivially
be extended beyond the standard amino acids. When the authors used but
the features based on their retention time index, they obtained a
negligible loss in accuracy compared to using all their original
features for unmodified peptides. Since this index was trained for
each dataset at hand, the algorithm became fully trainable for any
modified amino acid (Figure \ref{fig:elude}). The new version of {\sc
  ELUDE} was evaluated on peptides containing artifactual
modifications introduced as part of an MS protocol called COmbined
FRActional DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. The
results showed that {\sc ELUDE} yielded equally accurate predictions
for modified and unmodified peptides \cite{elude2}.

@@

Palmblad {\em et al.} showed that matching theoretical and observed
masses and retention times provides a simple, yet powerful method to
identify peptides in simple biological mixtures
\cite{palmblad2002prediction}. Strittmatter {\em et al.} used
retention time as a parameter in a discriminant function that
differentiated between correct and incorrect peptide identifications
\cite{Strittmatter2004}. Following this approach, they reported
$\sim$9\% more confident peptide identifications in a typical shotgun
experiment. Along the same lines, Klammer {\em et al.} demonstrated
that for specific experimental conditions, just filtering out peptides
with large deviations between observed and predicted retention time
leads to up to 50\% more identifications \cite{klammer2007improving}.

@@

The idea of identifying proteins based on entire peptides, rather than
peptide fragments, is not new. In fact, this was the principle
underlying peptide mass fingerprinting, one of the first methods used
to identify proteins via mass spectrometry. However, when introduced
in the 90's, mass fingerprinting could not be applied to unequivocally
identify proteins in complex biological samples. Consequently, shotgun
proteomics has gained popularity instead. Nevertheless, with great
developments in instrumentation leading to sub-ppm mass accuracy, the
possibility of using mass fingerprinting-like methods came back to
attention~\cite{iamt1}. The mentioned study showed how improved mass
accuracy and retention time predictions enable the use of unfragmented
ions even in the context of complex peptide mixtures.

@@

% Improve the design of SRM experiments 
In addition, predicted retention times are useful investigating the
elution properties of the peptides.  To illustrate this, in
\cite{elude1} the retention time of all the theoretical peptides from
an {\it in silico} digest of the human proteome were predicted using
{\sc ELUDE}. This exercise revealed a rather skewed distribution, with
a majority of the peptides eluting in only a small portion of the
chromatography run. This suggested that the linear gradient typically
used in liquid chromatography are sub-optimal, and that one should be
able to replace such gradients with nonlinear functions that produce
even distributions of the peptides across runs.

This hypothesis was tested by calculating and evaluating two types of
nonlinear gradients \cite{gradopt1}. The results showed that the new
gradients not only provided more uniform distributions of the
peptides, but also led to an increase in the number of peptide
identifications. In addition, they facilitated the identification of
different peptide species compared to the typical linear gradients. To
allow the easy calculation of nonlinear gradients by any MS
practitioner, a user-friendly graphical user interface was implemented 
\cite{gradopt2}.  Seen from a general perspective, this showed that
the gradient function is yet another parameter that researchers can
alter in their experiments, opening up a whole new range of potential
applications.

@@

The accuracy of the machine learning-based methods described in
section \ref{sec:ML} is still much lower than the methods based on
look-ups of previously registered retention times. However, the large
availability of datasets suggests that one possibility to increase the
accuracy of retention time predictions could rely on the use of
enhanced machine learning methods. Currently, there are a number of
such techniques that have not yet been applied to the problem of
predicting peptides' retention time. Here we will bring up one such
technique called pairwise Machine-Learned Ranking (MLR)
\cite{liu2009learning}, which we feel is particularly suitable for
this problem. MLR models can be trained to determine which of two data
points should be better ranked.

@@

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as big data, have brought into
attention so called deep learning techniques such as belief networks
using restricted multiple layered Boltzmann machines (RBM)
\cite{salakhutdinov2009deep} or deep neural networks. This follows the
observation that the performance of predictors is often enhanced if
one can structure the prediction into multiple layers. The drawback of
such models is that they require large training sets. Given the very
large sets of peptides with retention times that are assembled in
various repositories, such as Peptide Atlas
\cite{desiere2006peptideatlas} and PRIDE \cite{martens2005pride}, such
methods could probably be implemented for retention time prediction as
well.

@@

One way to determine a predictable separation technique's usefulness is
to estimate its ability to remove candidate peptides of any detected
feature. This can be measured by a false positive ratio (FPR), also
known as the method's fall-out. The FPR is of course a function of the
selected thresholds, but if we select a threshold where $\sim$95\% of
the known examples get correctly included, what FPR do we find at such
level?  Current retention time prediction algorithms have an FPR of
about $\frac{1}{5}$, meaning that it can remove about four out of five
candidate peptides for a given feature. This could be compared to the
FPR of about $\frac{1}{200}$ for predictions of high-resolution
isoelectric focusing experiments~\cite{branca2014hirief}. The
retention time prediction has to improve quite a bit to come to level
with the fall-out of isoelectric focusing, however, almost all mass
spectrometry-based proteomics use reversed phase chromatography, while
very few studies are using iso-electric focusing.

% An estimate of a FPR for MS1's would be great here.

Moreover, targeted proteomics experiments, such as Selected Reaction
Monitoring (SRM), are dependent on the predictability of a peptide's
behavior. In such experiments the retention times and the suitable
transitions are needed as input to the analysis. Frequently, such
information is derived from previous shotgun analysis. As the capacity
of targeted proteomics experiments expands by the introduction of the
more comprehensive data independent acquisition techniques
\cite{Venable2004} (a.k.a. SWATH analysis), such pre-identification of
peptide transitions and retention times threatens to become a limiting
factor. Our current methods to identify peptides from Data Independent
Acquisition are only as comprehensive as the previous shotgun analysis
would be. An attractive alternative is to predict retention time and
transition using {\em de novo} predictions from the peptides amino
acid sequences.  Such a system of identifications not dependent on
previous peptide identifications from shotgun proteomics experiments
would lead to a more direct experimental design and hopefully to more
comprehensive coverage. Thus {\em de novo} retention time predictions
have an important role to fill.

>>>>

Today a majority of the mass spectrometry (MS) protocols in use have a
high-throughput, generating extensive amounts of data. This abundance
of data has brought up a whole new set of analytical challenges,
making the development of efficient algorithms for data interpretation
a major area within proteomics research. A particularly difficult
challenge in this context is posed by the tremendous complexity of the
data.  Proteins display a wide range of physico-chemical properties
and are often found at very different amounts in complex samples,
spanning as much as 12 orders of magnitude in relative
abundance~\cite{Angel2012}.  Furthermore, a large proportion of
proteins are subjected to posttranslational
modifications~\cite{Lemeer2009}. To manage this diversity, increasing
efforts are made to extract additional information besides the
fragmentation spectra from each MS experiment.

An example of such additional information is represented by retention
times, {\em i.e.} the time points when peptides elute from the liquid
chromatography column.  The main purpose of liquid chromatography is
to reduce the number of peptides simultaneously injected into the mass
spectrometer. This reduction in complexity increases the probability
of each individual peptide to be selected for fragmentation, which in
its turn, increases the proteome coverage achieved in an experiment.
However, the separation process also in itself provides useful
information about the analytes. On the one hand, the instrument always
records the time when a peptide elutes from the chromatography column,
{\em i.e.}  the retention time. On the other hand, the retention time
can also be estimated by examining the sequence of the peptide. The
availability of both predicted and observed retention times has proven
to be a powerful feature, with several applications in MS-based
proteomics \cite{Palmblad2013}.

@@

At the moment, Reversed-Phase Liquid Chromatography (RPLC) is by far
the preferred method for separating peptides prior to mass
spectrometry analysis \cite{Xie201278}. This preference is mainly
because the method provides good resolution, gives reproducible
results, and most of all, is compatible with electrospray
ionization. In virtually all modern mass spectrometers the RPLC system
can be easily connected to the ionization source, such that the
peptides can be automatically transferred to mass spectrometry
analysis as they elute from the chromatography column.

@@

% say something about gradient and washing 
Because the proportion of solvent B determines the elution of
peptides, the scheme used to add solvent B is an essential aspect of
any RPLC experiment. Currently, the most commonly used approach is
based on {\em linear gradients}, where the increase in solvent B
follows a linear function. Most of the peptides elute before the
percentage of solvent B reaches 40\%, and thus gradients often go from
2 to 40\% B. A high increase in percentage B at the end of the run
ensures that all the remaining compounds are washed off from the
column.

@@

Liquid chromatography gives reproducible results. Except for small
shifts from run to run, the same peptide elutes at about the same time
when a sample is reanalyzed under identical experimental conditions
(Figure~\ref{fig:repr}). This consistency in elution time suggests
that peptides engage in well-defined interactions during separation,
which depend only on the experimental setup employed and the
properties of the peptides themselves. However, the properties of the
peptides are determined to a large extent by their amino-acid
sequence. As a consequence, given a chromatography setup, the sequence
of a peptide should provide virtually all the information needed to
predict its retention time.

@@

% The first method 
The first method to implement this idea used a set of 25 short
peptides together with their observed retention times to derive
retention coefficients for each of the amino acids and end groups
present in the sequences \cite{meek1980}. The coefficients were
calculated using a form of iterative regression and were estimated
using only the amino-acid composition of the peptides, without any
information about the position of each amino acid in the
sequence. However, because shuffled versions of the same peptide
sequence had previously been separated on a RPLC column, the author
pointed out that additional parameters would be required for
predicting the retention time of longer peptides.

@@

% It became available that additional information is needed; anc oclumn is important  examples 
Indeed, a subsequent study showed that even for short peptides,
distinct retention coefficients are obtained when an amino acid is
located at different positions in a sequence \cite{Houghten1987}. In
that context, the authors speculated that different retention
coefficients would be needed for each position in a peptide
sequence. Although the exact chemical mechanisms underlying that
observation are still not fully understood, a number of studies have
pinpointed individual factors that alter the retention coefficients of
amino acids. As an example, it has been shown that the retention
contribution of an amino acid is affected by the overall
hydrophobicity of the peptide \cite{Mant2006}, the number of positive
charges \cite{Mant2006} or presence of elements of secondary
structure \cite{Zhou1990}. In addition, modifications in
chromatography setup such as changing the composition of the mobile or
stationary phases also modify the elution pattern of the
peptides \cite{Browne1982, Guo1987, Gilar2010}.

@@

A limitation of index-based methods is that they are optimized for
predicting retention times for certain chromatography setups. Thus,
their prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}. Consequently, more advanced methods such as {\sc
  Elude} \cite{elude1,elude2} estimate retention time indices for each
dataset as hand. In these cases the indices are used to generate
features that are used as an input for final retention time
estimates. We will describe such methods more in detail in section
\ref{sec:ML}.

@@

The method was extended by representing each peptide by 1052 instead
of 20 features in a follow-up study~\cite{petritis2006improved}. The
new descriptors encoded the identity of each amino acid at each
position in a sequence, as well as the length and the hydrophobic
moment of the peptide. The new neural network was trained on more than
300,000 peptide sequences. To obtain such a large number of peptides
with known retention times, a strategy to normalize elution times was
developed. That strategy allowed pooling of peptides identified in
more than 12,000 shotgun experiments. The trained algorithm reported
better performance for the 1303 peptides on which it was evaluated,
than its predecessor.

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This requirement was a serious
limitation in the context of RPLC separation, where a broad variety of
chromatographic setups is commonly used across research
laboratories. To address that lack of flexibility, Klammer {\em et
  al.} proposed a predictor based on support vector machines (SVMs)
\cite{klammer2007improving}. Their method required fewer training
peptides and could thus be easily adapted to new chromatographic
conditions, but yielded lower prediction accuracy.  Subsequently, a
more elaborate SVM-based method, {\sc RTPredict}, that uses a
customized kernel function was proposed \cite{rtpredict,
  rtpredictImproved}. That approach provided an algorithm that
required as little as 50 peptides for training.

@@

Moruz {\em et al.}~\cite{elude1} developed a software package for
retention time prediction named {\sc Elude}. One of the main novelty
points of {\sc Elude} was the calculation of peptide features based on
both a general hydrophobicity index, as well as a retention index
trained for each dataset at hand. The features included properties
such as average hydrophobicity, the N- and C-termini's hydrophobicity,
the stretch of the five most and least hydrophobic amino acids,
helicity, and the number of occurrences of the 25\% most or least
hydrophobic amino acids. This scheme ensured that the algorithm could
be applied to different chromatography setups with minimal loss in
performance. For cases when little training data was available, {\sc
  Elude} implemented an additional workflow based on a library of
pre-trained retention models. This workflow required only a few tens
of training peptides to select and calibrate the most suitable model
from the library. Both the selection and calibration were performed
using statistical methods robust to outliers.

@@

To address the limitations of the previous studies, Moruz {\em et al.}
\cite{elude2} extended the first version of their algorithm {\sc
  ELUDE} to encompass modified peptides as well. The inclusion of such
peptides was straightforward to achieve once the authors observed that
many of the predictive features of their method were all functions of
the peptide's amino acid sequence and a retention index. As such
retention indices are estimated for each dataset at hand, at the
beginning of the training procedure, they can trivially be extended
beyond the standard amino acids. When the authors used but the
features based on their retention time index, they obtained a
negligible loss in accuracy compared to using all their original
features for unmodified peptides. Beacause this index was trained for
each dataset at hand, the algorithm became fully trainable for any
modified amino acid (Figure \ref{fig:elude}). The new version of {\sc
  ELUDE} was evaluated on peptides containing artifactual
modifications introduced as part of an MS protocol called COmbined
FRActional DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. The
results showed that {\sc ELUDE} yielded equally accurate predictions
for modified and unmodified peptides \cite{elude2}.

@@

Palmblad {\em et al.} showed that matching theoretical and observed
masses and retention times provides a simple, yet powerful method to
identify peptides in simple biological mixtures
\cite{palmblad2002prediction}. Strittmatter {\em et al.} used
retention time as a parameter in a discriminant function that
differentiated between correct and incorrect peptide identifications
\cite{Strittmatter2004}. Following that approach, they reported
$\sim$9\% more confident peptide identifications in a typical shotgun
experiment. Along the same lines, Klammer {\em et al.} demonstrated
that for specific experimental conditions, just filtering out peptides
with large deviations between observed and predicted retention time
leads to up to 50\% more identifications \cite{klammer2007improving}.

@@

The idea of identifying proteins based on entire peptides, rather than
peptide fragments, is not new. In fact, that strategy was the
principle underlying peptide mass fingerprinting, one of the first
methods used to identify proteins via mass spectrometry. However, when
introduced in the 90's, mass fingerprinting could not be applied to
unequivocally identify proteins in complex biological
samples. Consequently, shotgun proteomics has gained popularity
instead. Nevertheless, with great developments in instrumentation
leading to sub-ppm mass accuracy, the possibility of using mass
fingerprinting-like methods came back to attention~\cite{iamt1}. The
mentioned study showed how improved mass accuracy and retention time
predictions enable the use of unfragmented ions even in the context of
complex peptide mixtures.

@@

% Improve the design of SRM experiments 
In addition, predicted retention times are useful investigating the
elution properties of the peptides.  To illustrate this ability, in
\cite{elude1} the retention time of all the theoretical peptides from
an {\it in silico} digest of the human proteome were predicted using
{\sc ELUDE}. This exercise revealed a rather skewed distribution, with
a majority of the peptides eluting in only a small portion of the
chromatography run. This behavior suggested that the linear gradient
typically used in liquid chromatography are sub-optimal, and that one
should be able to replace such gradients with nonlinear functions that
produce even distributions of the peptides across runs.

This hypothesis was tested by calculating and evaluating two types of
nonlinear gradients \cite{gradopt1}. The results showed that the new
gradients not only provided more uniform distributions of the
peptides, but also led to an increase in the number of peptide
identifications. In addition, they facilitated the identification of
different peptide species compared to the typical linear gradients. To
allow the easy calculation of nonlinear gradients by any MS
practitioner, a user-friendly graphical user interface was implemented
\cite{gradopt2}.  Seen from a general perspective, these studies
showed that the gradient function is yet another parameter that
researchers can alter in their experiments, opening up a whole new
range of potential applications.

@@

The accuracy of the machine learning-based methods described in
section \ref{sec:ML} is still much lower than the methods based on
look-ups of previously registered retention times. However, the large
availability of datasets suggests that one possibility to increase the
accuracy of retention time predictions could rely on the use of
enhanced machine learning methods. Currently, there are a number of
such techniques that have not yet been applied to the problem of
predicting peptides' retention time. Here we will bring up one such
technique called pairwise Machine-Learned Ranking (MLR)
\cite{liu2009learning}, which we feel is particularly suitable for
that problem. MLR models can be trained to determine which of two data
points should be better ranked.

@@

It is also worth noting that the current trend of processing huge data
repositories, frequently referred to as big data, have brought into
attention so called deep learning techniques such as belief networks
using restricted multiple layered Boltzmann machines (RBM)
\cite{salakhutdinov2009deep} or deep neural networks. This remark
follows the observation that the performance of predictors is often
enhanced if one can structure the prediction into multiple layers. The
drawback of such models is that they require large training
sets. Given the very large sets of peptides with retention times that
are assembled in various repositories, such as Peptide Atlas
\cite{desiere2006peptideatlas} and PRIDE \cite{martens2005pride}, such
methods could probably be implemented for retention time prediction as
well.

@@

One way to measure a predictable separation technique's usefulness is
to estimate its ability to remove candidate peptides of any detected
feature. This property can be measured by a false positive ratio
(FPR), also known as the method's fall-out. The FPR is a function of
the selected thresholds, but if we select a threshold where $\sim$95\%
of the known examples get correctly included, what FPR do we find at
such level?  Current retention time prediction algorithms have an FPR
of about $\frac{1}{5}$, meaning that it can remove about four out of
five candidate peptides for a given feature. This rate could be
compared to the FPR of about $\frac{1}{200}$ for predictions of
high-resolution isoelectric focusing
experiments~\cite{branca2014hirief}. The retention time prediction has
to improve quite a bit to come to level with the fall-out of
isoelectric focusing, however, almost all mass spectrometry-based
proteomics use reversed phase chromatography, while very few studies
are using isoelectric focusing. As, an additional comparison is the
selection process based on the prediction of peptides' masses and
determination using mass spectrometry. A typical 10 ppm fully tryptic
search have on average $\sim 150$ candidate peptides per fragment
spectrum, out of the total of $\sim 4\cdot 10^6$ fully tryptic
peptides in the swissprot part of UniProt (including oxidation and
N-terminal acetylations), indicating that the FPR of the ms1-mass
determination/prediction is about $4\cdot 10^{-5}$. Neither
chromatographic retention time prediction, nor isoelectric focusing,
is currently near such performance.

Targeted proteomics experiments, such as Selected Reaction Monitoring
(SRM), are dependent on the predictability of a peptide's behavior. In
such experiments the retention times and the suitable transitions are
needed as input to the analysis. Frequently, such information is
derived from previous shotgun analysis. As the capacity of targeted
proteomics experiments expands by the introduction of the more
comprehensive data independent acquisition techniques
\cite{Venable2004} (a.k.a. SWATH analysis), such pre-identification of
peptide transitions and retention times threatens to become a limiting
factor. Our current methods to identify peptides from Data Independent
Acquisition are only as comprehensive as the previous shotgun analysis
would be. An attractive alternative is to predict retention time and
transition using {\em de novo} predictions from the peptides amino
acid sequences.  Such a system of identifications not dependent on
previous peptide identifications from shotgun proteomics experiments
would lead to a more direct experimental design and hopefully to more
comprehensive coverage. Thus {\em de novo} retention time predictions
have an important role to fill.
